---
title: "Minfi vs sesame comparison"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE, warning = FALSE)
```

## Setup

Loading packages

```{r, message=FALSE, warning=FALSE}
library(minfi)
library(sesame)
library(tidyverse)
```

Below is a custom function to go from idat files to beta values using **minfi**.

```{r}
idat_to_beta_values_noob <- function(...) {
  RGset <- read.metharray.exp(...)
  RGset@annotation = c(array = "IlluminaHumanMethylationEPIC", annotation = "ilm10b4.hg19")

  MSet.noob <- preprocessNoob(RGset, offset = 15, dyeCorr = TRUE, verbose = TRUE)

  ratioSet.noob <- ratioConvert(MSet.noob, what =  "both", keepCN = TRUE)
  beta.noob <- getBeta(ratioSet.noob)
  beta.noob
}
```

```{r}
idat_folder <- "~/Data/methylation/idat"
data_directory <- "~/Data/methylation/directory.csv"
```


## Minfi

Example of extraction of beta values for 1 sample using **minfi**.

```{r, message=FALSE}
minfi_data <- idat_to_beta_values_noob(
  base = idat_folder, 
  targets = data.frame(Basename = "200514030126_R06C01", 
                       stringsAsFactors = FALSE)
)
```

Plotting the distribution gives us a bimodal distribution.

```{r}
tibble(beta = minfi_data[, 1]) %>%
  ggplot(aes(beta)) +
  geom_histogram(bins = 100)
```

## Sessme

Example of extraction of beta values for 1 sample using **sesame**.

```{r}
sesame_data <- openSesame(file.path(idat_folder, "200514030126_R06C01"))
```

Plotting the distribution gives us a bimodal distribution. `NA` were dropped before plotting. 

```{r}
tibble(beta = sesame_data) %>%
  drop_na() %>%
  ggplot(aes(beta)) +
  geom_histogram(bins = 100)
```

## Visual comparison

The following function takes a vector of basenames and extracts the beta values using **minfi** and **sesame**. Sesame it set to perform pOOBAH masking which will show up as missing values.

```{r}
com_data <- function(x) {
  minfi_beta <- idat_to_beta_values_noob(base = idat_folder, 
                         targets = data.frame(Basename = x, 
                                              stringsAsFactors = FALSE))
  
  sesame_beta <- openSesame(file.path(idat_folder, x))
  
  data.frame(package = rep(c("minfi", "sesame"), c(length(minfi_beta), length(sesame_beta))),
             beta = c(minfi_beta[, 1], sesame_beta),
             site = c(rownames(minfi_beta), names(sesame_beta)),
             stringsAsFactors = FALSE)
}
```

A single sample is used and frequency polygons is created of the densities.

```{r}
sample_1 <- com_data(x = "200514030126_R06C01")

sample_1 %>%
  ggplot(aes(beta, color = package)) +
  geom_freqpoly(bins = 100)
```

## Investigate by plate and well id

The following function takes two arguments, plate and well and plots the density chart.

```{r}
com_plot <- function(plate, well) {
  read_csv(data_directory, 
         col_types = cols_only(
           plate = col_double(),
           well_position = col_character(),
           basename = col_character()
           )
         ) %>%
  filter(plate == .env$plate, well_position == well) %>%
  pull(basename) %>%
  com_data() %>%
  ggplot(aes(beta, color = package)) +
  geom_freqpoly(bins = 100) +
  labs(title = paste0(plate, "-", well))
}
```

```{r, message=FALSE, warning=FALSE}
com_plot(1378, "F01")
com_plot(1378, "H01")
com_plot(1385, "C01")
com_plot(1387, "H04")
com_plot(1464, "B01")
com_plot(1483, "A11")
```

## Comparing amount of masking by region

Following function take the arguments plate and well and returns the result of beta estimation. Works the same as `com_data` but input is plate and well instead of basename.

```{r}
com_region <- function(plate, well) {
  base_names <- read_csv(data_directory, 
         col_types = cols_only(
           plate = col_double(),
           well_position = col_character(),
           basename = col_character()
           )
         ) %>%
  filter(plate == .env$plate, well_position == well) %>%
  pull(basename)
  
  betas <- openSesame(file.path(idat_folder, base_names))
  
  tibble(site = names(betas),
         beta = betas,
         sample = paste(plate, well))
}
```

Following charts showchange the total count within each region compared to the count of masked cpgs.

```{r}
island_ref <- tibble(
island = Islands.UCSC@listData$Relation_to_Island,
rownames = Islands.UCSC@rownames
)
```


```{r}
ddd <- com_region(1378, "F01")

ddd %>%
  left_join(island_ref, by = c("site" = "rownames")) %>%
  group_by(island) %>%
  summarise(count = n(),
            is_na = sum(is.na(beta))) %>%
  pivot_longer(count:is_na) %>%
  ggplot(aes(island, value, fill = name)) +
  geom_col(position = "dodge")

ddd %>%
  left_join(island_ref, by = c("site" = "rownames")) %>%
  group_by(island) %>%
  summarise(pro_na = sum(is.na(beta)) / n()) %>%
  ggplot(aes(island, pro_na)) +
  geom_col()
```

## All the sample

Here we calculate the beta values for all the sample labeled as "tumor gland".

```{r all_data}
all_data <- read_csv(data_directory) %>%
  filter(description == "tumor gland") %>%
  select(plate, well_position) %>%
  pmap_dfr(~ com_region(.x, .y))
```

### Islands regions

Each sample is represented by a point in each column.

```{r, dependson="all_data"}
all_data %>%
   left_join(island_ref, by = c("site" = "rownames")) %>%
  group_by(sample, island) %>%
  summarise(pro_na = sum(is.na(beta)) / n()) %>%
  ggplot(aes(island, pro_na)) +
  geom_jitter(height = 0, width = 0.2)
```

A connected line chart shows that the variability of the procentage of masked points don't overlap.

```{r, dependson="all_data"}
all_data %>%
  mutate(island = Islands.UCSC@listData$Relation_to_Island[match(site, Islands.UCSC@rownames)]) %>%
  group_by(sample, island) %>%
  summarise(pro_na = sum(is.na(beta)) / n()) %>%
  ggplot(aes(island, pro_na)) +
  geom_line(aes(group = sample))
```

Below is a side by side comparrison between the number of sites within each region (on the left) and the procentages of cpg site for which at least of the the `r read_csv("~/Data/methylation/directory.csv") %>%
  filter(description == "tumor gland") %>% nrow()` sample have been masked.

```{r, dependson="all_data"}
fff <- all_data %>%
  group_by(site) %>%
  summarise(union_na = any(is.na(beta))) %>%
  mutate(island = Islands.UCSC@listData$Relation_to_Island[match(site, Islands.UCSC@rownames)]) 

fff %>%
  group_by(island) %>%
  summarise(count = n(),
            procent = mean(union_na)) %>%
  pivot_longer(count:procent) %>%
  mutate(name = factor(name, c("count", "procent"),
                       c("Total cpg count", "Percentage of cpgs"))) %>%
  ggplot(aes(island, value)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "how many cpg sites have at least 1 NA across all samples")
```

## Promotor regions

Below is same plots as in islands regions but with promotor regions instead.

```{r, dependson="all_data"}
all_data %>%
  mutate(island = Other@listData$Regulatory_Feature_Group[match(site, Other@rownames)]) %>%
  mutate(promotor = island == "Promoter_Associated") %>%
  group_by(sample, promotor) %>%
  summarise(pro_na = sum(is.na(beta)) / n()) %>%
  ggplot(aes(promotor, pro_na)) +
  geom_jitter(height = 0, width = 0.2)
```

```{r, dependson="all_data"}
fffpro <- all_data %>%
  group_by(site) %>%
  summarise(union_na = any(is.na(beta))) %>%
  mutate(island = Other@listData$Regulatory_Feature_Group[match(site, Other@rownames)]) %>%
  mutate(promotor = island == "Promoter_Associated") 

fffpro %>%
  group_by(island) %>%
  summarise(count = n(),
            procent = mean(union_na)) %>%
  pivot_longer(count:procent) %>%
  mutate(name = factor(name, c("count", "procent"),
                       c("Total cpg count", "Percentage of cpgs"))) %>%
  ggplot(aes(island, value)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~name, scales = "free_x") +
  labs(title = "how many cpg sites have at least 1 NA across all samples")
```



